---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
  download <- tempfile()
  download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",download)
  unzip(download)
  unlink(download)
}
```

## What is mean total number of steps taken per day?

```{r}
raw_data <- read.csv("activity.csv")
steps_day <-  aggregate(steps ~ date, raw_data, sum)
hist(steps_day$steps, main = "Steps per Day", xlab = "Steps", 
     ylab = "Frequency", ylim = c(0,40), col = "yellow")
```

### What are mean and median steps per day?
```{r}
step_mean <- mean(steps_day$steps)
step_med <-median(steps_day$steps)
cat("The mean number of steps per day is ", step_mean,".")
cat("The median number of steps per day is", step_med,".")
```

## What is the average daily activity pattern?
```{r}
interval_steps <- aggregate(steps ~ interval, raw_data, mean)

plot(interval_steps$interval, interval_steps$steps, type = "l", 
     col = "red", main = "Mean Steps per Interval", 
     xlab = "Interval", ylab = "Mean Steps", ylim = c(0,225))
```

###Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval <- which.max(interval_steps$steps)
max_index <-interval_steps[max_interval, ]
max_index <- c(max_index[1], max_index[2])
cat("The interval which contains the maximum number of steps and the number of steps is", ":")
max_index
```

## Imputing missing values
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
nrow(raw_data[is.na(raw_data$steps),])
```
## New Dataset replacing NAs with the mean per time period
```{r}
missing <- raw_data
for (i in 1:nrow(missing)){
  if (is.na(missing$steps[i])){
    missing$steps[i] <- interval_steps$steps[which(missing$interval[i]== interval_steps$interval)]
  }
}
```
## Histogram of new data with mean and median reported
```{r}
steps_day_missing <-  aggregate(steps ~ date, missing, sum)
hist(steps_day_missing$steps, main = "Steps per Day w/o Missing Data", 
     xlab = "Steps", ylab = "Frequency", ylim = c(0,40), col = "purple")
step_mean_missing <- mean(steps_day_missing$steps)
step_med_missing <-median(steps_day_missing$steps)
cat("The mean and median with values replaced are ", step_mean_missing, "and", 
    step_med_missing)
cat("Without replacing missing data, the mean and median daily steps were", step_mean, "and", step_med)
```
### The impact of replacing missing values with the corresponding interval mean is negligible.

## Weekend versus Weekday Activity
```{r}
missing$weekday <- weekdays(as.Date(missing$date, abbreviate = FALSE))

missing$whatday <- ifelse(missing$weekday == "Saturday" | 
        missing$weekday == "Sunday", "Weekend", "Weekday")

weekend <- subset(missing, missing$whatday == "Weekend", 
  select = c("steps", "date", "interval", "whatday"))

weekday <- subset(missing, missing$whatday == "Weekday", 
                  select = c("steps", "date", "interval", "whatday"))

weekend_steps <- aggregate(steps ~ interval, weekend, mean)
weekday_steps <- aggregate(steps ~ interval, weekday, mean)


par(mfcol = c(2,1), "mar" = c(3,2,2,1))

plot(weekend_steps$interval, weekend_steps$steps, type = "l", 
     col = "red", main = "Weekend Steps per Interval", 
     xlab = "Interval", ylab = "Mean Steps", ylim = c(0,225))

plot(weekday_steps$interval, weekday_steps$steps, type = "l", 
     col = "red", main = "Weekday Steps per Interval", 
     xlab = "Interval", ylab = "Mean Steps", ylim = c(0,225))
```
